---
title: 'Beyond Plot Labels: Hacking gganimate''s Frame Variables'
author: Karl Hailperin
date: '2019-07-16'
slug: []
categories: []
tags:
  - gganimate
  - datavis
description: ''
---

I was recently working on a fun project using `gganimate` and had one of those 
breakthroughs that made me think "This is what it feels like to steal fire from 
the gods!"[^1]

# An introduction to frame variables
One of [`gganimate`'s](https://cran.r-project.org/package=gganimate) cool
features is frame variables. They allow your labels to include data about the 
animation using `glue`-style syntax. For example:
```{r setup, message=FALSE}
library(tidyverse)
library(cowplot) # for theme_minimal_hgrid()
library(gganimate)
```

```{r basic_frame_var, warning=FALSE}
diamond_summary <- diamonds %>% 
  group_by(color, cut) %>% 
  summarise(ave_price = mean(price), max_price = max(price))

ggplot(diamond_summary, aes(color, ave_price)) + 
  geom_col() +
  theme_minimal_hgrid() +
  transition_states(cut, transition_length = 3, state_length = 1) +
  ggtitle("Going from {previous_state} to {next_state}")
```

We're using the frame variable `"{previous_state}"` to show that last value of 
`cut` we stopped at and `"{next_state}"` to show where we're going[^2]. 

But what if we wanted a title like "Max price for [{next_state}] is [price] with 
color: [color]"? This is the part where we get to play Prometheus. Thankfully,
`"{next_state}"` is a string we can pass to functions: 

```{r proof, warning=FALSE}
ggplot(diamond_summary, aes(color, ave_price)) + 
  geom_col() +
  theme_minimal_hgrid() +
  transition_states(cut, transition_length = 3, state_length = 1) +
  # cut is factor in diamonds, but is being coerced into a character
  ggtitle("next_state is a string: {is.character(next_state)}",
          subtitle = "next_state is a factor: {is.factor(next_state)}")
```

Note: Depending on the transition, type might be preserved instead of being 
coerced to character. Experiment to figure out how the transition you're using 
handles the data before writing any functions.


# Helper functions for frame variables
So now we just need a function to find which color has the highest price for a 
given `next_state`.

```{r helper_functions}
create_max_price_title <- function(frame_var){
  max_price_row <- diamond_summary %>% 
    filter(cut == frame_var) %>% 
    group_by(cut) %>% 
    filter(max_price == max(max_price))
  
  max_price <- pull(max_price_row, max_price)
  
  max_color <- pull(max_price_row, color)
  
  paste( "Max price for", frame_var,"is", max_price, "with color:", max_color)
}

```

Now we're good to go.
```{r anim_with_helper, warning=FALSE}
ggplot(diamond_summary, aes(color, ave_price)) + 
  geom_col() +
  theme_minimal_hgrid() +
  transition_states(cut, transition_length = 3, state_length = 1) +
  ggtitle("{create_max_price_title(next_state)}")
```
We did it!

# A couple of notes about `transition_reveal()` and `transition_time()`

First, unlike `transition_state()`, both `transition_reveal()` and 
`transition_time()` preserve the type of their frame variables. If your 
helper function relies on the fact that you are transitioning over a `Date`, 
for example, you don't have to convert it back from a character.

On the other hand, the frame variables for `transition_reveal()` and 
`transition_time()` often won't correspond to an actual value in your data the 
same way `next_state` does. `gganimate` creates artificial times evenly spaced 
across your data set and interpolates the appropriate values for your other 
variables[^3]. The frame variables are those artificial times. As result, 
your helper function will need to replace `filter(transition_col == frame_var)` 
with a step like `filter(transition_col <= frame_var)`. Depending on the 
helper function, you may want to use `slice()` or additional `filter()` calls to 
get just the most "recent" data.

[^1]: Minus the part where I get my liver repeatedly ripped out.
[^2]: Admittedly, it's a little awkward when they're the same because we're not 
in an in-between frame.
[^3]: This is an oversimplification of `gganimate`'s internals, not helped by 
the fact that I'm not an expert on tweening.
