---
title: Building Color Palette Proofs of Concept with purrr and ggplot2
author: Karl Hailperin
date: '2020-01-08'
tags:
  - dataviz
---
Recently, I raised the idea at work of using some new color palettes in our 
visualizations. I felt the colors we were using suffered from not being 
accessible to clients with color blindness and could benefit from being more
distinctive. My boss was open to the idea, but (understandably) didn't want
me to just start tinkering with our products. I needed to put together mockups 
of possible new palettes first. Fortunately, the tidyverse makes it fast to 
generate proofs of concept for several color options.

# Baseline plots
We want plots we can use to show off how different color palettes handle both
continuous and discrete data. A bar chart and a scatter plot seem like 
good ways to do that.

```{r setup, message = FALSE}
library(tidyverse)
library(scico)
library(patchwork) # to put both plots in the same image

theme_set(theme_minimal())
```

```{r bar_chart}
bar_chart <- ggplot(diamonds, aes(cut, fill = cut)) + 
  geom_bar() + 
  coord_flip() +
  # fill guide would just duplicate x-axis label and take up space
  guides(fill = FALSE) +
  labs(x = "Cut", y = "# of diamonds")

bar_chart
```

```{r scatter_plot, warning = FALSE}
scatter_plot <- ggplot(txhousing, aes(listings, sales, color = median)) +
  geom_point() +
  labs(x = "Monthly listings in county", 
       y = "# of sales",
       color = "Median sale price")

scatter_plot
```

# Palettes
We're going to use a few palettes from
[`viridis`](https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html) 
(now part of `ggplot2`) and 
[`scico`](https://www.data-imaginist.com/2018/scico-and-the-colour-conundrum/). 
Both packages prioritize palettes that are friendly to users with 
colorblindness and read well. Since this is for presenting to a higher-up, I've 
already narrowed it down to two or three palettes from each package. I pick ones
that I thought might be appealing and are fairly different from each other.

```{r palettes}
viridis_palettes <- c("plasma", "viridis")
# codes for plasma and viridis
viridis_pal_codes <- c("C", "D")

scico_palettes <- c("devon", "lajolla", "lapaz")
```

We're using a pair of vectors for `viridis` but not for `scico` 
because `ggplot2` uses letter codes to identify viridis palettes (e.g. `"C"` for 
`plasma`). However, we want to also have the palette name so that it can be 
used in plot titles and file names. If your manager says, "I really liked 
palette "B"; Let's use that one",  you don't want to be stuck scrambling to 
remember if it was "magma" or "inferno". `scico` just uses the name of the 
palette, so a vector of names is enough.

# Creating the Proofs of Concept
Since we're repeating the same core process (building a pair of plots) with 
minor tweaks (using a different color palette), we should write a function
that we can re-run with a new parameter for each palette. Adding a 
palette from `viridis` is different enough from adding one from `scico` that 
we'll actually want two functions, one for each package. However, both have the 
same core steps:
1. Add the palette to the baseline plots.
2. Combine the two plots into one with [`patchwork`](https://patchwork.data-imaginist.com/index.html) and add the 
name of the palette as a title.
3. Print out the new plot for review.
4. (Optional) Save the plot for later. I'm assuming that you're using a 
[project-oriented workflow](https://www.tidyverse.org/blog/2017/12/workflow-vs-script/) 
to manage your plot files.
```{r build_vir_poc}
build_vir_poc <- function(vir_pal_name, vir_pal_code, save_plot = FALSE){

# use the letter code part of the vector for the scale function
vir_bar <- bar_chart + scale_fill_viridis_d(option = vir_pal_code)

vir_scatter <- scatter_plot + scale_color_viridis_c(option = vir_pal_code)

# use the name part for the title
vir_patch <-  vir_bar/ vir_scatter + 
  plot_annotation(title = vir_pal_name)

print(vir_patch)

if (save_plot == TRUE){
  ggsave(filename = here::here(paste0("figs/", vir_pal_name, ".jpg")),
       plot = vir_patch)
}
}
```

```{r build_scico_poc}
build_scico_poc <- function(scico_pal, save_plot = FALSE){

scico_bar <- bar_chart + scale_fill_scico_d(palette = scico_pal)

scico_scatter <- scatter_plot + scale_color_scico(palette = scico_pal)
scico_patch <-  scico_bar/ scico_scatter + 
  plot_annotation(title = scico_pal)

print(scico_patch)

if (save_plot == TRUE){
  ggsave(filename = here::here(paste0("figs/", scico_pal, ".jpg")),
       plot = scico_patch)
}

}
```

# Scale up with `purrr::walk()`
To generate all those plots, we need to iterate the functions over our palettes.
We could do this with a `for loop`, but we can simplify things with 
`purrr::walk()`. `walk` is  version of
(`purrr::map()`)[https://jennybc.github.io/purrr-tutorial/ls01_map-name-position-shortcuts.html] 
that is designed for functions where we care about the side effects (e.g. 
creating a plot, writing to a file) instead of a traditional return value (e.g. 
the square root of an input). For `build_vir_poc`, we use a variant of `walk()`: 
`walk2()`. `walk2` iterates over a pair of vectors for when you have two 
parameters that you want to change. This is similar to a loop `for 
(i in 1:n){...}` where you use `i` multiple times inside of `{...}`.
```{r create_plots, warning = FALSE}
walk2(viridis_palettes, viridis_pal_codes, build_vir_poc)
walk(scico_palettes, build_scico_poc)
```

# Conclusion
While I like the `devon` colors, the fact it leads to a white bar on a white 
background is a deal breaker. I'd want to cut it even before presenting my 
findings. Otherwise, this is a good starting point for talking about 
alternatives to default colors. 

Once you've selected your color palettes, if you need to import them into 
another visualization tool, both `viridis` and `scico` make that 
straightforward.

```{r export_rgb}
# print 20 rgb codes for each palette
# plasma
viridisLite::viridis(20, option = "C")

# lapaz
scico(20, palette = "lapaz")

```

