---
title: Building Color Palette Proof of Concepts with purrr and ggplot2
author: Karl Hailperin
date: '2019-12-30'
slug: []
categories: []
tags:
  - dataviz
description: ''
---
Recently, I raised the idea at work of changing the color palettes we use for 
our dashboards. I felt the default ones we using suffered from not being 
accessible to clients with color blindness and could also benefit from being more
distinctive. My boss was open to the idea, but (understandably) wanted me to put 
together mockups of possible new palettes before we starting tinkering with our
product. Fortunately the tidyverse makes it fast to generate proof of concepts 
for several palettes.

# Baseline plots
We want plots we can use to show off how different color palettes handle 
continuous and discrete data. A bar chart and a scatter plot seem like 
good ways to do that.

```{r setup, echo = FALSE}
library(tidyverse)
library(patchwork) # to put both plots in the same image

theme_set(theme_minimal())
```

```{r bar_chart}
bar_chart <- ggplot(diamonds, aes(cut, fill = cut)) + 
  geom_bar() + 
  coord_flip() +
  # fill guide would just duplicate x-axis label and take up space
  guides(fill = FALSE) +
  labs(x = "Cut", y = "# of diamonds")

bar_chart
```

```{r scatter_plot}
scatter_plot <- ggplot(txhousing, aes(listings, sales, color = median)) +
  geom_point() +
  labs(x = "Monthly listings in county", 
       y = "# of sales",
       color = "Median sale price")

scatter_plot
```

```{r patchwork}
patch <- bar_chart / scatter_plot

patch
```

# Palettes
